Climate variability indices for ecological and crop models in R: the `climatrends` package

### 2022-02-10 inst/paper/paper.Rmd
travis

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Magne Neby
    affiliation: "3"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
  - name: Department of Forest and Field Sciences, Inland Norway University of Applied Sciences, 2480 Koppang, Norway
    index: 3
citation_author: de Sousa et. al.
date: "A preprint"
year: 2020
bibliography: paper.bib
csl: citationstyle.csl
output:
  pdf_document:
    keep_tex: false
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

```{r get data, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("climatrends")
library("PlackettLuce")
load("cbean.rda")
```

\pagenumbering{gobble}

>Kauê de Sousa^1,2[*]^, Jacob van Etten^2^, Svein Øivind Solberg^1^  
^1^ Department of Agricultural Sciences, Inland Norway University of Applied Sciences, 2318 Hamar, Norway  
^2^ Bioversity International, 00054 Maccarese, Rome, Italy    
^[*]^ Correspondence should be addressed to: <kaue.desousa@inn.no>

\

@@

Abiotic factors play an important role in most ecological and crop systems that depend on certain levels of temperature, light and precipitation (and their interplay) to initiate important physiological events [@PlantEcology]. In the walk of climate change, understanding how these factors drive the physiological processes is a key approach to provide recommendations for adaptation and biodiversity conservation. The package `climatrends` aims to provide the methods in R [@RCoreTeam] to compute precipitation and temperature indices that serve as input for climate and crop models [@vanEtten2019; @Kehel2016], trends in climate change [@Aguilar2005; @deSousa2018] and applied ecology [@Prentice1992; @YLiu2018].

# Methods and features

## Implementation

Six main functions are provided (Table 1), `crop_sensitive()`, `ETo()`, `GDD()`, `late_frost()`, `rainfall()` and `temperature()` with a default method for numeric 'vector' and additional methods implemented via the package `methods` [@RCoreTeam] for classes 'matrix' (or array), 'data.frame', and 'sf' (of geometry POINT or POLYGON) [@sf]. The last two methods are designed to fetch data from cloud sources, until now from the packages `nasapower` [@Sparks2018] and `chirps` [@chirps].

@@

These functions started as a set of scripts to compute indices in citizen science trials. In these trials. Aiming to capture the environmental variation across different sites, which can differ since each data-point generally have a different starting day and duration, the arguments `day.one` and `span` are vectorised and may be variable across data-points. For time series analysis, where fixed periods are defined across many locations, the indices can be adjusted with the argument `last.day` linked to the argument `day.one`.

@@

## Crop-related indices

@@

The crop sensitive indices available in **climatrends** are described in Table 3. These indices were previously used in crop models to project the impacts of climate change on crop yield [@Challinor2016; @Trnka2014]. Each index has a default temperature threshold(s) which can be adjusted by using the arguments `*.threshold`. Where the `*` means the index. For example, to change the defaults for hts_max (high temperature stress), a vector with the temperature thresholds is passed through the argument `hts_max.thresholds`.

@@

# compute the number of days required to accumulate
# gdd from planting date to maturity
gdd <- GDD(modis, 
           day.one = cbean$planting_date, 
           degree.days = 900, 
           return.as = "ndays")

# add gdd to the cbean data and take the average 
# of gdd per season
cbean %<>%  
  mutate(gdd = gdd$gdd) %>% 
  group_by(season) %>% 
  mutate(gdds = as.integer(mean(gdd)))
```

To compute the temperature indices we use the array with tempeture data, the vector with planting dates, and the seasonal averaged degree-days passed as a vector using the argument `span`. `temperature()` concatenates the data from the given `day.one` to the given `span` and compute the indices for each row.

@@

```{r pltree_bean, message=FALSE, echo=TRUE, eval=FALSE}
# compute the temperature indices from planting date to the 
# number of days required to accumulate the gdd in each season
temp <- temperature(modis, 
                    day.one = cbean$planting_date, 
                    span = cbean$gdds)

# combine the indices with the main data
cbean <- cbind(cbean, temp)

@@

Across-season distribution of maxNT captured for each sample plot in this experiment is shown in Fig. 1b. The data has a bimodal distribution which is reflected in the splitting value (18.7 $^\circ C$) for the Plackett-Luce trees in Fig. 1c. The upper node splits with 49 summer days (SU). We can interpret these results as that differences in growing performance of common beans is led by a considerable amount of diurnal temperature above a warmer threshold of 30 $^\circ C$ (in this case >70% of the growing days) and warmer nights (> 18.7 $^\circ C$).

```{r fig_cbean, fig.cap="Fig. 1. Application of climatrends functions to support the analysis of a citizen-science data testing 11 common bean varieties in Nicaragua. (A) Days required to reach 900 growing-degree days from planting date calculated using the function GDD(). (B) Maximum night temperature (°C) distributed across seasons computed using the function temperature(). (C) Plackett-Luce Tree showing the probability of one common bean variety has to win against the others (axys X) in three different nodes splitted with the summer days (day temperature > 30 °C) and maximum night temperature (°C). Note: the first season (primera, Pr) spans from May to August, the second (postrera, Po) from September to October, and the third (apante, Ap) from November to January.", out.width = '80%', echo=FALSE, fig.align="center"}
knitr::include_graphics("cbean.png")
```

@@

We randomly selected 100 points in hexagonal within the coordinates 7$^\circ$ and 17$^\circ$ W, and 59 $^\circ$ and 63 $^\circ$ N, that comprises Norway and Sweden before the Arctic Circle. We compute the temperature indices from 2000-01-01 to 2019-12-31 using the function `temperature()` with the method for objects of class 'sf'. The temperature data is fetched from from the NASA Langley Research Center POWER Project funded through the NASA Earth Science Directorate Applied Science Program (<https://power.larc.nasa.gov/>), using the R package `nasapower` [@Sparks2018].

@@

The package can support the integration with other datasets as they become available in `R` via API client packages. Also new indices related to the physiology of crops could be implemented.

@@

# Data availability statement

To explore the latest functionalities of `climatrends`, please check the package's updates at CRAN (https://cran.r-project.org/package=climatrends).

>>>>

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Magne Neby
    affiliation: "3"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Systems Transformation Group, Bioversity International, Montpellier, France
    index: 2
  - name: Department of Forest and Field Sciences, Inland Norway University of Applied Sciences, 2480 Koppang, Norway
    index: 3
citation_author: de Sousa et. al.
year: 2022
bibliography: paper.bib
output: rticles::joss_article
journal: JOSS
# output:
#   pdf_document:
#     keep_tex: false
# header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

```{r get data, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("climatrends")
library("PlackettLuce")
```

@@

Abiotic factors play an important role in most ecological and crop systems that depend on certain levels of temperature, light and precipitation to initiate important physiological events [@PlantEcology]. Understanding how these factors drive the physiological processes is a key approach to provide recommendations for adaptation and biodiversity conservation in applied ecology studies. The package `climatrends` aims to provide the methods in R [@RCoreTeam] to compute precipitation and temperature indices that serve as input for climate and crop models [@vanEtten2019; @Kehel2016], trends in climate change [@Aguilar2005; @deSousa2018] and applied ecology [@Prentice1992; @YLiu2018].

# Implementation

Six main functions are provided (Table 1), with a default method for numeric 'vector' and additional methods implemented via the package `methods` [@RCoreTeam] for classes 'matrix' (or array), 'data.frame', and 'sf' (of geometry POINT or POLYGON) [@sf]. The last two methods are designed to fetch data from cloud sources, currently from the packages `nasapower` [@Sparks2018] and `chirps` [@chirps].

@@

These functions started as a set of scripts to compute indices from on-farm testing sites following a citizen science approach [@vanEtten2019]. Aiming to capture the environmental variation across different sites, which can differ as each on-farm trial generally have a different starting day and duration, the arguments `day.one` and `span` are vectorised and may be used to indicate the starting date for each data-point and the duration of the timespan to be considered for the computation of the indices. For time series analysis, fixed periods can be adjusted with the argument `last.day` linked to the argument `day.one`.

@@

## Crop ecology indices

@@

The crop ecology indices available in **climatrends** are described in Table 3. These indices were previously used in crop models to project the impacts of climate change on crop yield [@Challinor2016; @Trnka2014]. Each index has a default temperature threshold(s) which can be adjusted by using the arguments `*.threshold`. Where the `*` means the index. For example, to change the defaults for hts_max (high temperature stress), a vector with the temperature thresholds is passed through the argument `hts_max.thresholds`.

@@

# number of days required to accumulate gdd from planting date to maturity
gdd <- GDD(modis, 
           day.one = cbean$planting_date, 
           degree.days = 900, 
           return.as = "ndays")

# gdd to the cbean data and take the average gdd per season
cbean %<>%  
  mutate(gdd = gdd$gdd) %>% 
  group_by(season) %>% 
  mutate(gdds = as.integer(mean(gdd)))
```

To compute the temperature indices we use the array with temperature data, the vector with planting dates, and the seasonal averaged degree-days passed as a vector using the argument `span`. The function `temperature()` concatenates the data from the given `day.one` to the given `span` and computes the indices for each row.

@@

```{r pltree_bean, message=FALSE, echo=TRUE, eval=FALSE}
# temperature indices from planting date to the 
# number of days required to accumulate the gdd in each season
temp <- temperature(modis, 
                    day.one = cbean$planting_date, 
                    span = cbean$gdds)

cbean <- cbind(cbean, temp)

@@

Across-season distribution of maxNT captured for each sample plot in this experiment is shown in Fig. 1b. The data has a bimodal distribution which is reflected in the splitting value (18.7 $^\circ C$) for the Plackett-Luce trees in Fig. 1c. The upper node splits with 49 summer days (SU). We can interpret these results as that the on-farm performance of common beans varieties is led by heat accumulation of diurnal temperature above 30 $^\circ C$ (in this case >70% of the growing days) and warmer nights (> 18.7 $^\circ C$).

```{r fig_cbean, fig.cap="Fig. 1. Application of climatrends functions to support the analysis of on-farm trial data. (A) Days required to reach 900 growing-degree days from planting date calculated using the function GDD(). (B) Maximum night temperature (°C) distributed across seasons computed using the function temperature(). (C) Plackett-Luce Tree showing the probability that a given variety to outperform the other varieties (axys X) in three different nodes splitted with the summer days (day temperature > 30 °C) and maximum night temperature (°C). Note: the first season (primera, Pr) spans from May to August, the second (postrera, Po) from September to October, and the third (apante, Ap) from November to January.", out.width = '80%', echo=FALSE, fig.align="center"}
knitr::include_graphics("cbean.png")
```

@@

We randomly selected 100 points in hexagonal within the coordinates 7$^\circ$ and 17$^\circ$ W, and 59 $^\circ$ and 63 $^\circ$ N, that comprises Norway and Sweden before the Arctic Circle. We compute the temperature indices from 2000-01-01 to 2019-12-31 using the function `temperature()` with the method for objects of class 'sf'. The temperature data is fetched from the NASA Langley Research Center POWER Project funded through the NASA Earth Science Directorate Applied Science Program (<https://power.larc.nasa.gov/>), using the R package `nasapower` [@Sparks2018].

@@

The package can support the integration with other datasets as they become available in `R` via API client packages. Also new indices related to the physiology of crops could be implemented. To explore the latest functionalities of `climatrends`, please check the package's updates at CRAN (https://cran.r-project.org/package=climatrends).

@@
